LayeredTypes -- Combining dependent and independent type systems

### 2023-03-10 paper/types2023.tex
References
Small pass

% Some of our commands for this guide.
%
\authorrunning{Abelt and Fonseca}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

@@

@@

In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.

@@

 To illustrate this in a small example, consider the program in \Cref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line. In this short snippet, there are multiple concerns to be addressed. Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types. Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that does not exceed the list length. Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution. Note that the same standard library functions (\lstinline|createFD|, \lstinline|readLines|, etc...) may be needed in other programs with different type systems requirements.

@@

These properties can be verified in two major ways.

All features from all type systems can be combined in single, monolithic type system, containing all the complexity that all the different type systems entail. A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow such distinction, leading to confusing and unintelligible error messages~\cite{wits-error-messages}.

@@

@@

We propose a second, more principled alternative: \LayeredTypes. In \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies. The missing layers can be added over time, if they are found to be relevant.

We want to note an important distinction to Gradual Typing~\cite{gradual,gradual-objects}: Gradual Typing allows to only provide partial typing information in one system, but does not allow to choose different sets of types to verify. It allows to blend between having no types (0) and having types (1). Our proposal supports multiple type systems, with different combinations among them.

We evaluated our approach in a prototype language. Users can provide their own implementations for verification layers and define dependencies between them. In \autoref{lst:code_after} we see a set of annotations that can be added to the base code of \autoref{lst:code_before} to tackle the issues described. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

@@

\input{chapters/acknowledgements.tex}

\bibliographystyle{plain}
\bibliography{bibliography}

\end{document}
%------------------------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{literature}

>>>>

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

@@

\authorrunning{Authorrunning}

@@

\section{Challenges in Type Systems}
\label{sec:challenges-in-type-systems}

Right now, to choose a particular type system, developers have to change programming languages, which might introduce interoperability problems with other components of their system. Furthermore, authors of new programming languages have to write extensive standard libraries in order for their language to be adopted by the masses.
 Most commonly used programming languages such as Java, C++ or Haskell embed a single type system that checks, either statically or dynamically, that functions and variables are used in a compatible way.

 To use more sophisticated type systems such as Liquid Types, developers either port their programs to a different programming language, or rely on compiler plugins to add another type-checking phase, like in LiquidHaskell\cite{LiquidHaskell} or LiquidJava\cite{LiquidJava}. However, it is not clear how multiple type systems would interact in the same programming language.

 \paragraph{Problem statement:} How can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.

@@

 To illustrate this in a small example, consider the program in \Cref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line. In this short snippet, there are multiple concerns to be addressed. Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types. Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that exceeds the list length. Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution.

@@

Optimally, these properties should be verified independently from each other wherever possible. For example tracking the state of a resource can be done independently of it's specific type, so in case that for whatever reason this check fails we still want to be able to independently verify whether or not the state behaves as we intend it.

@@

A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow us to define distinct predicates to be verified independently. In such cases we have to merge all relevant properties into one joint predicate which can make it hard to identify the specific issue when the verification fails.

@@

To tackle these issues, we present \LayeredTypes. In our approach we view different type systems as indepent layers that can be verified in an incremental way. Layers can either verify properties in a complete stand-alone matter, or they can define dependencies to other layers which allow to build more complex type systems.

We want to note an important distinction to Gradual Typing\cite{gradual}\cite{gradual-objects}: While the latter is concerned with only providing partial typing information and making it gradually more precise, our approach does not focus on the information provided to the verifier, but rather on \textit{how} this information.

In our work we present a python-based demonstration framework that implements this layered verification approach. Users can provide their own implementations for verification layers and define dependencies between them. In \autoref{lst:code_after} we see a set of annotations that can be added to the base code of \autoref{lst:code_before} to tackle the issues described. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

@@

%------------------------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{literature}
